Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2023-05-05 paper.Rmd
revised introduction

---
title: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
type: ARTICLE TEMPLATE
author:
  - name: Nicholas Spyrison
    affiliation: a
    orcid: 0000-0002-8417-0212
    email: spyrison@gmail.com
  - name: Dianne Cook
    affiliation: a
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
  - name: Przemyslaw Biecek
    affiliation: b
    orcid: 0000-0001-8423-1823
    email: przemyslaw.biecek@gmail.com
affiliation:
  - num: a
    address: |
      Monash University, Melbourne, Australia
  - num: b
    address: |
      Warsaw University of Technology and University, Warsaw, Poland
abstract: |
  The increased predictive power of nonlinear models comes at the cost of interpretability of its terms. This trade-off has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models use predictors to arrive at a prediction.  _Local explanations_, which provide a point estimate of the linear variable importance in the vicinity of one observation is one XAI method. These can be considered to be linear projections and can be further explored to understand the interactions between variables used to make predictions across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any observation and illustrate with examples from categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) response models. The methods are implemented in the __R__ package __cheem__, available on CRAN.
keywords: | 
  explainable artificial intelligence; nonlinear model interpretability; visual analytics; local explanations; grand tour; radial tour
bibliography: paper.bib
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

@@

<!-- Paper structure -->
The remainder of this paper is organized as follows. Section \ref{sec:explanations} covers the background of the local explanation and the traditional visuals produced. Section \ref{sec:tour} explains the tours and particularly the radial manual tour. Section \ref{sec:cheemviewer} discusses the visual layout in the graphical user interface and how it facilitates analysis, data preprocessing, and package infrastructure. Illustrations are provided in Section \ref{sec:casestudies} for a range of supervised learning tasks with categorical and quantitative outputs. Section \ref{sec:cheemdiscussion} concludes with a summary of the insights gained. The methods are implemented in the __R__ package __cheem__. Example data sets are provided to illustrate usage for classification and regression tasks.

@@

<!-- Reminder of local explanation -->
Consider a highly nonlinear model. It can be hard to determine whether small changes in a variable's value will make a class prediction change group or identify which variables contribute to an extreme residual. Local explanations shed light on these situations by approximating linear variable importance in the vicinity of a single observation.

<!-- Taxonomy of local explanations -->
A comprehensive summary of the taxonomy and literature on explanation techniques is provided in Figure 6 by @arrieta_explainable_2020. It includes a large number of model-specific explanations such as deepLIFT [@shrikumar_not_2016; @shrikumar_learning_2017], a popular recursive method for estimating importance in neural networks. There are fewer model-agnostic explanations, of which LIME, [@ribeiro_why_2016] SHAP, [@lundberg_unified_2017], and their variants are popular.

<!-- Uses of local explanations -->
These observation-level explanations are used in various ways depending on the data. In image classification, where pixels correspond to predictors, saliency maps overlay or offset a heatmap indicating important pixels [@simonyan_deep_2014]. For observation, pixels corresponding to snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis highlights the sentiment and magnitude of influential words [@vanni_textual_2018]. In the case of numeric regression, they are used to explain variable additive contributions from the model intercept to the observation's prediction [@ribeiro_why_2016].

<!-- SHAP -->
SHaply Additive exPlanations (SHAP) quantifies the variable contributions of one observation by examining the effect of other variables on the predictions. The SHAP explanation refers to @shapley_value_1953's method to evaluate an individual's contribution in cooperative games by assessing this player's performance in the presence or absence of other players. @strumbelj_efficient_2010 introduced SHAP for local explanations in ML models. The attribution of variable importance depends on the sequence of the included variables. The SHAP values are the mean contributions over different variable sequences. The approach is related to partial dependence plots [@molnar_interpretable_2020], used to explain the effect of a variable by predicting the response for a range of values on this variable after fixing the value of all other variables to their mean. Though partial dependence plots are a global approximation of the variable importance, while SHAP is specific to one observation. It could also be considered similar to examining the coefficients from all subsets regression, as described in @wickham_visualizing_2015 help to understand the relative importance of each variable in the context of all other candidate variables.

@@

<!-- Fifa example -->
Following the use case _Explanatory Model Analysis_ [@biecek_explanatory_2021], FIFA data is used to  illustrate SHAP. Consider soccer data from the FIFA 2020 season [@leone_fifa_2020]. There are 5000 observations of 9 skill measures (after aggregating highly correlated variables). A random forest model is fit regressing players' wages [2020 Euros] from their skill measurements. The SHAP values are compared for a star offensive player (L. Messi) and defensive player (V. van Dijk). The results are displayed in Figure \@ref(fig:shapdistrbd). A difference in the attribution of the variable importance across the two positions of the players can be expected. This would be interpreted as how a player's salary depends on this combination of skills. Panel (a) is a modified breakdown plot [@gosiewska_ibreakdown_2019] where three sequences of variables are presented, so the two observations can be more easily compared. The magnitude of the contributions depends on the sequence in which they appear. Panel (b) shows the differences in the player's median values of over 25 such sequences.<!-- Segue --> In summary, these plots highlight how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this observation, two players with different positions receive different profiles of variable importance to explain the prediction of their wages.

<!-- tree SHAP -->
For the application, we use _tree SHAP_, a variant of SHAP that enjoys a lower computational complexity [@lundberg_consistent_2018]. Instead of aggregating over sequences of the variables, tree SHAP calculates observation-level variable importance by exploring the structure of the decision trees. Tree SHAP is only compatible with tree-based models; random forests are used for illustration. The following section will use normalized SHAP values as a projection basis (call this the _attribution projection_) that will be used to explore the sensitivity of the variable contributions.

@@

To explore the local explanations, coordinated views [@roberts_state_2007] [also known as ensemble graphics, @unwin_ensemble_2018] are provided in the _cheem viewer_ application. There are two primary plots: the __global view__ to give the context of all of the SHAP values and the __radial tour view__ to explore the local explanations with user-controlled rotation. There are numerous user inputs, including variable selection for the radial tour and observation selection for making comparisons. There are different plots used for the categorical and quantitative responses. Figures \@ref(fig:classificationcase) and \@ref(fig:regressioncase) are screenshots showing the cheem viewer for the two primary tasks: classification (categorical response) and regression (quantitative response).

@@

<!-- Approximations of the spaces, PC1:2 -->
The visualization is composed of the first two principal components of the data (left) and the attribution (middle) spaces. These single 2D projections will not reveal all of the structure of higher-dimensional space, but they are helpful visual summaries. In addition, a plot of the observed against predicted response values is also provided (Figures \@ref(fig:classificationcase)b, \@ref(fig:regressioncase)a) to help identify observations poorly predicted by the model. <!-- Interactions --> For classification tasks, misclassified observations are circled in red. Linked brushing between the plots is provided, and a tabular display of selected points helps to facilitate exploration of the spaces and the model (shown in Figures \@ref(fig:regressioncase)d).

@@

<!-- Local explanations as par coords -->
The local explanations for all observations are normalized (sum of squares equals 1), and thus, the relative importance of variables can be compared across all observations. These are depicted as vertical parallel coordinate plots [@ocagne_coordonnees_1885] on the basis biplot [@gabriel_biplot_1971]. 1D biplot displays the values of the current basis as bars. The parallel coordinate overlays lines connecting one observation's variable attribution (Figures \@ref(fig:classificationcase)e and \@ref(fig:regressioncase)e). The attribution projections of the PI and CI are shown as dashed and dotted lines. From this plot, the range and density of the importance across all observations can be interpreted. For classification, one would look at differences between groups on any variable. For example, Figure \@ref(fig:classificationcase)e suggests that `bl` is important for distinguishing the green class from the other two. For regression, one might generally observe which variables have low values for all observations (not important). For example, `BMI` and `pwr` in Figure \@ref(fig:regressioncase)e, have a range of high and low values (e.g., `off`, `def`), suggesting they are important for some observations and not important for others.

@@

<!-- Tour animation -->
The attribution projection of the PI is the initial 1D basis in a radial tour, displayed as a density plot for a categorical response (Figure \@ref(fig:classificationcase)f) and as scatterplots for a quantitative response (Figure \@ref(fig:regressioncase)f). The PI and CI are indicated by vertical dashed and dotted lines. The radial tour varies the contribution of the selected variable between 0 and 1. This is viewed as an animation of the projections from many intermediate bases. Doing so tests the sensitivity of structure (class separation or strength of relationship) to the variable's contribution. For classification, if the separation between classes diminishes when the variable contribution is reduced, this suggests that the variable is important for class separation. For regression, if the relationship scatterplot weakens when the variable contribution is reduced, indicating that the variable is important for accurately predicting the response.

@@

Selecting a misclassified observation as PI and a correctly classified point nearby in data space as CI makes it easier to examine the variables most responsible for the error. The global view (Figure \@ref(fig:classificationcase)c) displays the model confusion matrix. The radial tour is 1D and displays as density where color indicates class. An animation slider enables users to vary the contribution of variables to explore the sensitivity of the separation to that variable.

```{r classificationcase, echo=F, out.width = "100%", fig.align="center", fig.cap = "Overview of the cheem viewer for classification tasks (categorical outputs). Global view inputs, (a), set the PI, CI, and color statistic. Global view, (b) PC1 by PC2 approximations of the data- and attribution-space. (c) prediction by observed $y$ (visual of the confusion matrix for classification tasks). Points are colored by predicted class, and red circles indicate misclassified observations. Radial tour inputs (d) select variables to include and which variable is changed in the tour. (e) shows a parallel coordinate display of the distribution of the variable attributions while bars depict contribution for the current basis. The black bar is the variable being changed in the radial tour. Panel (f) is the resulting data projection indicated as density in the classification case."}
knitr::include_graphics("./figures/app_classification.PNG")
```

@@

Selecting an inaccurately predicted observation as PI and an accurately predicted observation with similar variable values as CI is a helpful way to understand how the model is failing or not. The global view (Figure \@ref(fig:regressioncase)a) shows a scatterplot of the observed vs predicted values, which should exhibit a strong relationship if the model is a good fit. The points can be colored by a statistic, residual, a measure of outlyingness (log Mahalanobis distance), or correlation to aid in understanding the structure identified in these spaces.

@@

```{r regressioncase, echo=F, out.width="100%", fig.cap="Overview of the cheem viewer for regression tasks (quantitative outputs) and illustration of interactive variables. Panel (a) PCA of the data- and attributions- spaces and the (b) residual plot, predictions by observed values. Four selected points are highlighted in the PC spaces and tabularly displayed. Coloring on a statistic (c) highlights structure organized in the attribution space. Interactive tabular display (d) populates when observations are selected. Contribution of the 1D basis affecting the horizontal position (e) parallel coordinate display of the variable attribution from all observations, and horizontal bars show the contribution to the current basis. Regression projection (f) uses the same horizontal projection and fixes the vertical positions to the observed $y$ and residuals (middle and right)."}
knitr::include_graphics("./figures/app_regression_interactions.PNG")
```

@@

<!-- Reactive vs exploration interactions -->
The application has several reactive inputs that affect the data used, aesthetic display, and tour manipulation. These reactive inputs make the software flexible and extensible (Figure \@ref(fig:classificationcase)a & d). The application also has more exploratory interactions to help link points across displays, reveal structures found in different spaces, and access the original data.

<!-- Exploratory interactions -->
A tooltip displays observation number/name and classification information while the cursor hovers over a point. Linked brushing allows the selection of points (left click and drag) where those points will be highlighted across plots (Figure \@ref(fig:classificationcase)a & b). The information corresponding to the selected points is populated on a dynamic table (Figure \@ref(fig:classificationcase)d). These interactions aid exploration of the spaces and, finally, the identification of primary and comparison observations.

@@

Figure \@ref(fig:casepenguins) shows plots from the cheem viewer for exploring the random forest model on the penguins data. Panel (a) shows the global view, and panel (b) shows several 1D projections generated with the radial tour. Penguin 243, a Gentoo (purple), is the PI because it has been misclassified as a Chinstrap (orange).

@@

Figure \@ref(fig:casepenguinsblfl) confirms that flipper length (`fl`) is vital for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted, and the PI can be seen to be closer to the Chinstrap group in these two variables, mainly because it has an unusually low value of flipper length relative to other Gentoos. From this view, it makes sense that it is a hard observation to account for, as decision trees can only partition only vertical and horizontal lines.

@@

<!-- Dark misclassified as milk, description -->
This type of exploration is shown in Figure \@ref(fig:casechocolates), where a chocolate labeled dark but predicted to be milk is chosen as the PI (observation 22). It is compared with a CI that is a correctly classified dark chocolate (observation 7). The PCA plot and the tree SHAP PCA plots (a) show a big difference between the two chocolate types but with confusion for a handful of observations. The misclassifications are more apparent in the observed vs predicted plot and can be seen to be mistaken in both ways: milk to dark and dark to milk.

@@

<!-- segue and selection for chocolates inverse case -->
It would also be interesting to explore an inverse misclassification. In this case, a milk chocolate is selected while it was misclassified as a dark chocolate. Chocolate 84 is just this case and is compared with a correctly predicted milk chocolate (observation 71). The corresponding global view and radial tour frames are shown in Figure \@ref(fig:casechocolatesinverse).

@@

<!-- Data description -->
The 2020 season FIFA data [@leone_fifa_2020; @biecek_dalex_2018] contains many skill measurements of soccer/football players and wage information. Nine higher-level skill groupings were identified and aggregated from highly correlated variables. A random forest model is fit from these predictors, regressing player wages [2020 euros]. The model was fit from 5000 observations before being thinned to 500 players to mitigate occlusion and render time. Continuing from the exploration in Section \@ref(sec:explanations), we are interested to see the difference in attribution based on the exogenous player position. That is, the model should be able to use multiple linear profiles to better predict the wages from different field positions of players despite not having this information. A leading offensive fielder (L. Messi) is compared with a top defensive fielder (V. van Dijk). The same observations were used in Figure \@ref(fig:shapdistrbd).

@@

Figure \@ref(fig:casefifa) tests the support of the local explanation. Offensive and reaction skills (`off` and `rct`) are both crucial to explaining a star offensive player. If either of them were rotated out, the other would be rotated into the frame, maintaining a far-right position. However, increasing the contribution of a variable with low importance would rotate both variables out of the frame.

@@

Figure \@ref(fig:caseames) selects the house sale 74, a sizable under prediction with an enormous Lot Area contribution. The CI has a similar predicted price though the prediction was accurate and gives almost no attribution to lot size. The attribution projection places observations with high Living Areas to the right. The contribution of Living Area contrasts the contribution of this variable. As the contribution of Lot Area decreases, the predictive power decreases for the PI, while the CI remains stationary. This large importance in the Living Area is relatively uncommon. Boosting tree models may be more resilient to such an under-prediction as they would up-weighting this residual and force its inclusion in the final model.

>>>>

---
title: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
type: ARTICLE TEMPLATE
author:
  - name: Nicholas Spyrison
    affiliation: a
    orcid: 0000-0002-8417-0212
    email: spyrison@gmail.com
  - name: Dianne Cook
    affiliation: a
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
  - name: Przemyslaw Biecek
    affiliation: b
    orcid: 0000-0001-8423-1823
    email: przemyslaw.biecek@gmail.com
affiliation:
  - num: a
    address: |
      Monash University, Melbourne, Australia
  - num: b
    address: |
      Warsaw University of Technology and University, Warsaw, Poland
abstract: |
  The increased predictive power of nonlinear models comes at the cost of interpretability of its terms. This trade-off has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models use predictors to arrive at a prediction.  "Local explanations", which provide a point estimate of the linear variable importance in the vicinity of one observation, are one XAI method. These can be considered to be linear projections and can be further explored to understand the interactions between variables used to make predictions across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any observation and illustrate with examples from categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) response models. The methods are implemented in the __R__ package __cheem__, available on CRAN.
keywords: | 
  explainable artificial intelligence; nonlinear model interpretability; visual analytics; local explanations; grand tour; radial tour
bibliography: paper.bib
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

@@

<!-- Paper structure -->
The remainder of this paper is organized as follows. Section \ref{sec:explanations} covers the background of the local explanation and the traditional visuals produced. Section \ref{sec:tour} explains the tours and particularly the radial manual tour. Section \ref{sec:cheemviewer} discusses the visual layout in the graphical user interface and how it facilitates analysis, data pre-processing, and package infrastructure. Illustrations are provided in Section \ref{sec:casestudies} for a range of supervised learning tasks with categorical and quantitative response variables. These show how the local explanations can be used to get an overview of the model's use of predictors, and to investigate errors in the model predictions. Section \ref{sec:cheemdiscussion} concludes with a summary of the insights gained. The methods are implemented in the __R__ package __cheem__.

@@

<!-- Reminder of local explanation -->
Consider a highly nonlinear model. It can be hard to determine whether small changes in a variable's value will make a class prediction change group or identify which variables contribute to an extreme residual. Local explanations shed light on these situations by estimate linear variable importance in the vicinity of a single observation. There are many approaches for calculating local explanations.

<!-- Taxonomy of local explanations -->
A comprehensive summary of the taxonomy of currently available methods is provided in Figure 6 by @arrieta_explainable_2020. It includes a large number of model-specific explanations such as deepLIFT [@shrikumar_not_2016; @shrikumar_learning_2017], a popular recursive method for estimating importance in neural networks. There are fewer model-agnostic methods, of which LIME, [@ribeiro_why_2016] SHaply Additive exPlanations (SHAP), [@lundberg_unified_2017], are popular.

<!-- Uses of local explanations -->
These observation-level explanations are used in various ways depending on the data. In image classification, where pixels correspond to predictors, saliency maps overlay or offset a heatmap to indicate important pixels [@simonyan_deep_2014]. For example, pixels corresponding to snow may be highlighted as important contributors when distinguishing if a picture contains a coyote or husky. In text analysis, word-level contextual sentiment analysis highlights the sentiment and magnitude of influential words [@vanni_textual_2018]. In the case of numeric regression, they are used to explain additive contributions of variables from the model intercept to the observation's prediction [@ribeiro_why_2016].

<!-- SHAP -->
We will be focusing on SHAP values in this this paper, but the approach is applicable for any methods of calculating the local explanations. SHAP calculates the variable contributions of one observation by examining the effect of other variables on the predictions. The SHAP explanation refers to @shapley_value_1953's method to evaluate an individual's contribution in cooperative games by assessing this player's performance in the presence or absence of other players. @strumbelj_efficient_2010 introduced SHAP for local explanations in machine learning models. The attribution of variable importance depends on the sequence being the included variables. The SHAP value for an observation are the average contributions of each variable, computed for different sequences, or permutations, of the variables. The approach is related to partial dependence plots (see for example @molnar_interpretable_2020), used to explain the effect of a variable by predicting the response for a range of values on this variable after fixing the value of all other variables to their mean. Though partial dependence plots are a global approximation of the variable importance, while SHAP is specific to one observation. It could also be considered similar to examining the coefficients from all subsets regression, as described in @wickham_visualizing_2015 which help to understand the relative importance of each variable in the context of all other candidate variables.

@@

<!-- Fifa example -->
We use 2020 season FIFA data [@leone_fifa_2020] to illustrate SHAP following the procedures described in @biecek_explanatory_2021. There are 5000 observations of nine predictor variables measuring players' skills and one response variable, wages (in euros). A random forest model is fit regressing players' wages on the skill variables. In this illustration in Figure \ref{fig:shapdistrbd} the SHAP values are compared for a star offensive player (L. Messi) and a prominent defensive player (V. van Dijk). What we are interested to know is how the skill variables locally contribute to the wage prediction of each player. A difference in the attribution of the variable importance across the two positions of the players can be expected. This would be interpreted as how a player's salary depends on which combination of skills. Panel (a) is a version of a breakdown plot [@gosiewska_ibreakdown_2019] where just three sequences of variables are shown. Here we can see how the variable contribution can change depending on sequence, relative to both players. (Note that the order of the variables is different in each plot, because they have been sorted by biggest average contribution across both players.) For all sequences, and for both players `reaction` has the strongest contribution, with perhaps more importance for the defensive player. Then it differs by player: for Messi `offense` and `movement` have the strongest contributions, and for van Dijk it is `defense` and `power`, regardless of the variable sequence.

Panel (b) shows the differences in the player's median values (large dots) for 25 such sequences (tick marks).  We can see that the wage predictions for the two players come from different combinations of skills sets, as might be expected for players who's value on the team depends on their offensive or defensive prowess. It is also interesting to see from the distribution of values across the different sequence of variables, that there is some multimodality. For example, look at the SAHP values for `reaction` for Messi, and in some sequences reaction has a much lower contribution than others. This suggests that other variables (`offense`, `movement` probably) can substitute for `reaction` in the wage prediction.

In summary, these plots highlight how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this observation, two players with different positions receive different profiles of variable importance to explain the prediction of their wages.

<!-- tree SHAP -->
For the application, we use _tree SHAP_, a variant of SHAP that enjoys a lower computational complexity [@lundberg_consistent_2018]. Instead of aggregating over sequences of the variables, tree SHAP calculates observation-level variable importance by exploring the structure of the decision trees. Tree SHAP is only compatible with tree-based models. so random forests are used for illustration. <!--The following section will use normalized SHAP values as a projection basis (call this the _attribution projection_) that will be used to explore the sensitivity of the variable contributions.-->

@@

To explore the local explanations, coordinated views [@roberts_state_2007] [also known as ensemble graphics, @unwin_ensemble_2018] are provided in the _cheem viewer_ application. There are two primary plots: the __global view__ to give the context of all of the SHAP values and the __radial tour view__ to explore the local explanations with user-controlled rotation. There are numerous user inputs, including variable selection for the radial tour and observation selection for making comparisons. There are different plots used for the categorical and quantitative responses. Figures \ref{fig:classificationcase} and \ref{fig:regressioncase} are screenshots showing the cheem viewer for the two primary tasks: classification (categorical response) and regression (quantitative response).

@@

<!-- Approximations of the spaces, PC1:2 -->
The visualization is composed of the first two principal components of the data (left) and the attribution (middle) spaces. These single 2D projections will not reveal all of the structure of higher-dimensional space, but they are helpful visual summaries. In addition, a plot of the observed against predicted response values is also provided (Figures \ref{fig:classificationcase}b, \ref{fig:regressioncase}a) to help identify observations poorly predicted by the model. <!-- Interactions --> For classification tasks, misclassified observations are circled in red. Linked brushing between the plots is provided, and a tabular display of selected points helps to facilitate exploration of the spaces and the model (shown in Figures \ref{fig:regressioncase}d).

@@

<!-- Local explanations as par coords -->
The local explanations for all observations are normalized (sum of squares equals 1), and thus, the relative importance of variables can be compared across all observations. These are depicted as vertical parallel coordinate plots [@ocagne_coordonnees_1885] on the basis biplot [@gabriel_biplot_1971]. 1D biplot displays the values of the current basis as bars. The parallel coordinate overlays lines connecting one observation's variable attribution (Figures \ref{fig:classificationcase}e and \ref{fig:regressioncase}e). The attribution projections of the PI and CI are shown as dashed and dotted lines. From this plot, the range and density of the importance across all observations can be interpreted. For classification, one would look at differences between groups on any variable. For example, Figure \ref{fig:classificationcase}e suggests that `bl` is important for distinguishing the green class from the other two. For regression, one might generally observe which variables have low values for all observations (not important). For example, `BMI` and `pwr` in Figure \ref{fig:regressioncase}e, have a range of high and low values (e.g., `off`, `def`), suggesting they are important for some observations and not important for others.

@@

<!-- Tour animation -->
The attribution projection of the PI is the initial 1D basis in a radial tour, displayed as a density plot for a categorical response (Figure \ref{fig:classificationcase}f) and as scatterplots for a quantitative response (Figure \ref{fig:regressioncase}f). The PI and CI are indicated by vertical dashed and dotted lines. The radial tour varies the contribution of the selected variable between 0 and 1. This is viewed as an animation of the projections from many intermediate bases. Doing so tests the sensitivity of structure (class separation or strength of relationship) to the variable's contribution. For classification, if the separation between classes diminishes when the variable contribution is reduced, this suggests that the variable is important for class separation. For regression, if the relationship scatterplot weakens when the variable contribution is reduced, indicating that the variable is important for accurately predicting the response.

@@

Selecting a misclassified observation as PI and a correctly classified point nearby in data space as CI makes it easier to examine the variables most responsible for the error. The global view (Figure \ref{fig:classificationcase}c) displays the model confusion matrix. The radial tour is 1D and displays as density where color indicates class. An animation slider enables users to vary the contribution of variables to explore the sensitivity of the separation to that variable.

```{r classificationcase, echo=F, out.width = "100%", fig.align="center", fig.cap = "Overview of the cheem viewer for classification tasks (categorical response). Global view inputs, (a), set the PI, CI, and color statistic. Global view, (b) PC1 by PC2 approximations of the data- and attribution-space. (c) prediction by observed $y$ (visual of the confusion matrix for classification tasks). Points are colored by predicted class, and red circles indicate misclassified observations. Radial tour inputs (d) select variables to include and which variable is changed in the tour. (e) shows a parallel coordinate display of the distribution of the variable attributions while bars depict contribution for the current basis. The black bar is the variable being changed in the radial tour. Panel (f) is the resulting data projection indicated as density in the classification case."}
knitr::include_graphics("./figures/app_classification.PNG")
```

@@

Selecting an inaccurately predicted observation as PI and an accurately predicted observation with similar variable values as CI is a helpful way to understand how the model is failing or not. The global view (Figure \ref{fig:regressioncase}a) shows a scatterplot of the observed vs predicted values, which should exhibit a strong relationship if the model is a good fit. The points can be colored by a statistic, residual, a measure of outlyingness (log Mahalanobis distance), or correlation to aid in understanding the structure identified in these spaces.

@@

```{r regressioncase, echo=F, out.width="100%", fig.cap="Overview of the cheem viewer for regression tasks (quantitative response) and illustration of interactive variables. Panel (a) PCA of the data- and attributions- spaces and the (b) residual plot, predictions by observed values. Four selected points are highlighted in the PC spaces and tabularly displayed. Coloring on a statistic (c) highlights structure organized in the attribution space. Interactive tabular display (d) populates when observations are selected. Contribution of the 1D basis affecting the horizontal position (e) parallel coordinate display of the variable attribution from all observations, and horizontal bars show the contribution to the current basis. Regression projection (f) uses the same horizontal projection and fixes the vertical positions to the observed $y$ and residuals (middle and right)."}
knitr::include_graphics("./figures/app_regression_interactions.PNG")
```

@@

<!-- Reactive vs exploration interactions -->
The application has several reactive inputs that affect the data used, aesthetic display, and tour manipulation. These reactive inputs make the software flexible and extensible (Figure \ref{fig:classificationcase}a & d). The application also has more exploratory interactions to help link points across displays, reveal structures found in different spaces, and access the original data.

<!-- Exploratory interactions -->
A tooltip displays observation number/name and classification information while the cursor hovers over a point. Linked brushing allows the selection of points (left click and drag) where those points will be highlighted across plots (Figure \ref{fig:classificationcase}a & b). The information corresponding to the selected points is populated on a dynamic table (Figure \ref{fig:classificationcase}d). These interactions aid exploration of the spaces and, finally, the identification of primary and comparison observations.

@@

Figure \ref{fig:casepenguins} shows plots from the cheem viewer for exploring the random forest model on the penguins data. Panel (a) shows the global view, and panel (b) shows several 1D projections generated with the radial tour. Penguin 243, a Gentoo (purple), is the PI because it has been misclassified as a Chinstrap (orange).

@@

Figure \ref{fig:casepenguinsblfl} confirms that flipper length (`fl`) is vital for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted, and the PI can be seen to be closer to the Chinstrap group in these two variables, mainly because it has an unusually low value of flipper length relative to other Gentoos. From this view, it makes sense that it is a hard observation to account for, as decision trees can only partition only vertical and horizontal lines.

@@

<!-- Dark misclassified as milk, description -->
This type of exploration is shown in Figure \ref{fig:casechocolates}, where a chocolate labeled dark but predicted to be milk is chosen as the PI (observation 22). It is compared with a CI that is a correctly classified dark chocolate (observation 7). The PCA plot and the tree SHAP PCA plots (a) show a big difference between the two chocolate types but with confusion for a handful of observations. The misclassifications are more apparent in the observed vs predicted plot and can be seen to be mistaken in both ways: milk to dark and dark to milk.

@@

<!-- segue and selection for chocolates inverse case -->
It would also be interesting to explore an inverse misclassification. In this case, a milk chocolate is selected while it was misclassified as a dark chocolate. Chocolate 84 is just this case and is compared with a correctly predicted milk chocolate (observation 71). The corresponding global view and radial tour frames are shown in Figure \ref{fig:casechocolatesinverse}.

@@

<!-- Data description -->
The 2020 season FIFA data [@leone_fifa_2020; @biecek_dalex_2018] contains many skill measurements of soccer/football players and wage information. Nine higher-level skill groupings were identified and aggregated from highly correlated variables. A random forest model is fit from these predictors, regressing player wages [2020 euros]. The model was fit from 5000 observations before being thinned to 500 players to mitigate occlusion and render time. Continuing from the exploration in Section \ref{sec:explanations), we are interested to see the difference in attribution based on the exogenous player position. That is, the model should be able to use multiple linear profiles to better predict the wages from different field positions of players despite not having this information. A leading offensive fielder (L. Messi) is compared with a top defensive fielder (V. van Dijk). The same observations were used in Figure \ref{fig:shapdistrbd}.

@@

Figure \ref{fig:casefifa} tests the support of the local explanation. Offensive and reaction skills (`off` and `rct`) are both crucial to explaining a star offensive player. If either of them were rotated out, the other would be rotated into the frame, maintaining a far-right position. However, increasing the contribution of a variable with low importance would rotate both variables out of the frame.

@@

Figure \ref{fig:caseames} selects the house sale 74, a sizable under prediction with an enormous Lot Area contribution. The CI has a similar predicted price though the prediction was accurate and gives almost no attribution to lot size. The attribution projection places observations with high Living Areas to the right. The contribution of Living Area contrasts the contribution of this variable. As the contribution of Lot Area decreases, the predictive power decreases for the PI, while the CI remains stationary. This large importance in the Living Area is relatively uncommon. Boosting tree models may be more resilient to such an under-prediction as they would up-weighting this residual and force its inclusion in the final model.
